---
title: "SITA Analysis with EISA and DESeq"
output: html_notebook
---
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir="/Users/Pomato/mrc/project/sita/")
```


## Packages
```{r packages}
suppressPackageStartupMessages({
  library(eisaR)
  library(ggplot2)
  library(dplyr)
  library(glue)
  library(DESeq2)
  library(biomaRt)
})
```

## Functions
```{r}
add_ensembl_symbol <- function (table) {
  genes <- row.names(table)
  
  if (grepl("ENSG", genes[1], fixed=TRUE)) {
    ensemblDataset <- "hsapiens_gene_ensembl"
    symbol <- "hgnc_symbol"
  } else if (grepl("ENSMUSG", genes[1], fixed=TRUE)) {
    ensemblDataset <- "mmusculus_gene_ensembl"
    symbol <- "mgi_symbol"
  }
  
  mart <- useDataset(ensemblDataset, useMart("ENSEMBL_MART_ENSEMBL", host="www.ensembl.org"))
  geneList <- getBM(filters="ensembl_gene_id",
                    attributes=c("ensembl_gene_id", symbol),
                    values=genes,
                    mart=mart)
  
  row.names(geneList) <- geneList[, 1]
  geneList[, 1] <- NULL
  
  table$geneSymbol <- geneList[, 1][match(rownames(table), rownames(geneList))]
  newTable <- table
  
  return(newTable)
}
```

## Load count data
Takes raw count tables of introns and exons obtained using the package QuasR. Details on how to obtain count tables for this analysis [here](https://rdrr.io/github/fmicompbio/eisaR/f/vignettes/eisaR.Rmd).
```{r load.data}
gse <- "GSE120807"
selectConditions <- c("unstimulated", "^LPS") # control vs. treatment

exon <- read.table(glue("processed/{gse}_ExonicCounts.txt"), header=TRUE, sep="\t", row.names=1)
intron <- read.table(glue("processed/{gse}_IntronicCounts.txt"), header=TRUE, sep="\t", row.names=1)

exon <- exon %>% dplyr::select(matches(selectConditions)) %>% as.matrix()
intron <- intron %>% dplyr::select(matches(selectConditions)) %>% as.matrix()

head(exon)
```

## EISA
Run Exon Intron Split Analysis (EISA) on the raw intron and exon count tables for a certain GSE using . Outputs a MA plot that shows stress-induced transcriptional attenuation (SITA), a global downregulation in genes upon stress induction.
```{r eisa}
# Filter for genes which have ≥ 1 exon and intron
shared <- intersect(rownames(exon), rownames(intron))
exonsh <- exon[shared, ]
intronsh <- intron[shared, ]

message("No. of genes with ≥ 1 exon and intron: ", nrow(exonsh))

# Checks
allsh <- exonsh + intronsh
fracIn <- colSums(intronsh)/colSums(allsh)
summary(fracIn)

# Format and select conditions for each sample
cond <- gsub("_rep[0-9]*.*", "", colnames(exonsh))
conditions <- unique(cond)

stopifnot(all(colnames(exonsh) == colnames(intronsh)))

res_eisar <- runEISA(cntEx=exonsh, cntIn=intronsh,
                     cond=cond,
                     method=NULL,
                     modelSamples=TRUE,
                     geneSelection="filterByExpr",
                     statFramework="QLF",
                     effects="predFC",
                     pscnt=2,
                     recalcNormFactAfterFilt=TRUE,
                     recalcLibSizeAfterFilt=FALSE)

# Flip logFC
res_eisar$tab.ExIn$logFC <- -(res_eisar$tab.ExIn$logFC)

# Add gene symbol
res_eisar$tab.ExIn <- add_ensembl_symbol(res_eisar$tab.ExIn)

# EISA MA plot
MAplot <- ggplot(res_eisar$tab.ExIn, aes(x=logCPM, y=logFC)) +
  geom_point(color="lightgrey") +
  geom_point(data=res_eisar$tab.ExIn %>% filter(FDR < 0.05),
             color="red") +
  labs(title=glue("{gse}: {conditions[1]} vs. {conditions[2]}"),
       caption=glue("UP={nrow(filter(res_eisar$tab.ExIn, FDR < 0.05 & logFC > 0))}, DOWN={nrow(filter(res_eisar$tab.ExIn, FDR < 0.05 & logFC < 0))}")) +
  theme_bw() +
  theme(plot.title=element_text(size=15, face="bold"),
        plot.caption=element_text(size=15))

MAplot

# Save output
png(glue("processed/{gse}_eisaMAplot_{conditions[1]}.{conditions[2]}.png"))
print(MAplot)
dev.off()

deGenes <- res_eisar$tab.ExIn %>% arrange(FDR, -logFC)
write.table(deGenes, file=glue("./processed/{gse}_eisaDE_{conditions[1]}.{conditions[2]}.txt"), sep="\t", row.names=TRUE, col.names=TRUE)
```

## DESeq2
Using the raw counts table for exons, the significant differentially expressed (DE) genes are determined with the package DESeq2.
```{r deseq}
# Paired samples
paired <- FALSE

# Make column metadata
cond <- gsub("_rep[0-9]*.*", "", colnames(exon))
conditions <- unique(cond)

if (paired) {
  colData <- data.frame(row.names=colnames(exon),
                        replicate=factor(gsub(".rep", "", colnames(exon)),
                                         levels=seq.int(1:(ncol(exon) / 2))),
                        condition=factor(cond,
                                         levels=conditions))

  dataset <- DESeqDataSetFromMatrix(countData=exon,
                                  colData=colData,
                                  design=~ replicate + condition)
} else {
  colData <- data.frame(row.names=colnames(exon),
                        condition=factor(cond, levels=conditions))

  dataset <- DESeqDataSetFromMatrix(countData=exon,
                                    colData=colData,
                                    design=~ condition)
}

# Set reference level as control
dataset$condition <- relevel(dataset$condition, ref=conditions[1])

dataset <- dataset[rowSums(counts(dataset)) >= 10, ] # pre-filter genes with no reads

## Run DESeq
dds <- DESeq(dataset)

res <- results(dds, contrast=c('condition', conditions), alpha=0.05)
res <- res[complete.cases(res), ] # remove any rows with NA

summary(res)
message("No. of significant DE genes (FDR < 0.05): ", sum(res$padj < 0.05, na.rm=TRUE))

# Add gene symbol
res <- add_ensembl_symbol(res)

# Save DESeq results table to output
deGenes <- as.data.frame(res) %>% arrange(padj, desc(log2FoldChange)) # order by adjusted p-value and FC
write.table(deGenes,
            file=glue("processed/{gse}_DESeq_{conditions[1]}.{conditions[2]}.txt"),
            row.names=TRUE, col.names=TRUE, sep="\t")

## Visualisation
resLFC <- lfcShrink(dds, coef=tail(resultsNames(dds), n=1), type="apeglm")

# DESeq2 MA plot
png(glue("processed/{gse}_DESeqMAplot_{conditions[1]}.{conditions[2]}.png"))
print(DESeq2::plotMA(resLFC, main=glue("{gse}: {conditions[1]} vs. {conditions[2]}
                                           UP={nrow(filter(deGenes, padj < 0.05 & log2FoldChange > 0))}, DOWN={nrow(filter(deGenes, padj < 0.05 & log2FoldChange < 0))}")))
dev.off()
```

